A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-25 paper/fascd.tex
text in intro, and cleanup near end

FIXME organization of paper

@@

\section{Basic CD iterations} \label{sec:cditers}

Associated to a constraint decomposition (CD; equations \eqref{eq:subspacedecomp}--\eqref{eq:constraintrestrictionsum}) are certain iterative methods \cite{Tai2003,Xu1992}.  Let $0\le \alpha \le 1$ be the \emph{damping parameter}.  Algorithms \ref{alg:basiccd-add} and \ref{alg:basiccd-mult} below, discussed next, solve smaller VI problems over each subset $\cK_i$.  Each Algorithm replaces $u \in \cK$ with a new iterate $w\in\cK$ which should be closer to $u^* \in \cK$ solving \eqref{eq:vi}.

@@

Compared specifically to the $V(1,0)$ cycle presented as Algorithm 4.7 in \cite{GraeserKornhuber2009}, one might say informally that \pr{fascd-vcycle} is a $V(1,1)$ cycle.  However, its structure relates directly to the basic CD iterations (Sections \ref{sec:cditers}, \ref{sec:cdmultilevel}) as well.  In unilateral cases it is precise to say that one iteration of \pr{fascd-vcycle} is an application of \pr{cd-mult} over CD \eqref{eq:fe:downwardsuminclusion}, using the defect constraint meaning of the downward constraint sets $\mathcal{D}^j$, followed by an application of \pr{icd-tele} over the incomplete CD \eqref{eq:fe:upwardsuminclusion}.  FIXME: idea is that once you are using defect constraints you no longer depend on the literal availability of $\Pi_i$, so incomplete CD suffices, e.g.~even for \pr{cd-mult}

@@

FIXME stopping criteria for smoother

FIXME F-cycles in Algorithm \ref{alg:fascd-fcycle}

\begin{pseudofloat}[ht]
\begin{pseudo}
\pr{fascd-fcycle} \\
FIXME
\end{pseudo}
\caption{FIXME}
\label{alg:fascd-fcycle}
\end{pseudofloat}

@@

% A BRIDGE TOO FAR:  \section{Results for a nonlocal variational inequality}

@@

@@

\appendix
\section{Reduction of FASCD to FAS multigrid for PDEs} \label{app:reductions}

\pr{fascd-vcycle} (Algorithm \ref{alg:fascd}) in the main text generalizes existing multilevel algorithms.  Here we show that it reduces to the FAS multigrid V-cycle for nonlinear PDEs \cite{Trottenbergetal2001}, which itself generalizes linear multigrid, when all inequality constraints are removed.

Suppose we remove the inequality constraints from the finest-level VI problem \eqref{eq:fe:vi} by replacing the finest-level obstacles $\underline{\gamma}^J$ and $\overline{\gamma}^J$ with very negative and positive constants, respectively, thus not constraining the solution $u^J$ or any possible iterate.  The admissible set $\mathcal{K}^J$ of \eqref{eq:fe:vi} is then determined only by the Dirichlet boundary conditions.  Let $\mathcal{V}_0^J \subset \mathcal{V}^J$ denote the linear subspace with zero values on $\partial_D\Omega$.  By replacing the (now) arbitrary element $v-u^J\in\mathcal{V}_0^J$ in \eqref{eq:fe:vi} with ``$v$'' we have the finest-level problem
\begin{equation}
\ip{f^J(u^J)}{v} = \ip{\ell^J}{v} \qquad \text{for all } v\in \mathcal{V}_0^J, \label{eq:app:fas:pde}
\end{equation}
the FE discretization of strong-form PDE $f(u)=\ell$.  The downward $\mathcal{D}^j$ and upward $\mathcal{U}^j$ constraint sets can also be enlarged so that the corrections $y^j$ and $z^j$ in \pr{fascd-vcycle} are never constrained.\footnote{Formally we might do this by choosing constant LDCs with $\underline{\chi}^j \ll 0 \ll \overline{\chi}^j$, but so that ordering property \eqref{eq:fe:chiordering} still holds, and also with large gaps so that $\underline{\phi}^j = \underline{\chi}^j - \underline{\chi}^{j-1} \ll 0 \ll \overline{\phi}^j = \overline{\chi}^j - \overline{\chi}^{j-1}$.}  Thus the corrections $y^j$ and $z^j$ are unconstrained, and lines 2--3 and 5--8 can be removed from Algorithm \ref{alg:fascd}.

Now the downward/upward VIs \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi} are also unconstrained, and can be rewritten as weak forms $\ip{f^j(w^j + y^j)}{v} = \ip{\ell^j}{v}$ and $\ip{f^j(w^j + z^j)}{v} = \ip{\ell^j}{v}$, respectively, for all $v\in \mathcal{V}_0^j$.  However, we may simply store the unconstrained, approximate solution $u^j=w^j+y^j$, instead of $w^j$ and $y^j$ separately, and similarly for the upward solution $u^j=w^j+z^j$.  The smoothers and the coarsest-level solver can be regarded as unconstrained, in-place modifications of $u^j$, so we denote them simply as ``$\text{\pr{smooth}}(\ell^j;u^j)$.''  Noting line 11 of Algorithm \ref{alg:fascd} now says $u^{j-1}=\iR u^j$, after down-smoothing the coarsened source functional computed by line 12 can be rewritten as
\begin{equation}
\ell^{j-1} = f^{j-1}\left(u^{j-1}\right) + R\left(\ell^j-f^j(u^j)\right). \label{eq:app:fas:levelsource}
\end{equation}

Going upward we must be careful that only the coarse correction, and not the coarse solution, is prolonged.\footnote{Compare Remark 5.3.9 in \cite{Trottenbergetal2001}, a warning to novices and everyone else.}  After up-smoothing on the $j-1$ level the coarse iterate is $u^{j-1}$, but at this point $u^j=w^j + y^j$ contains the result of down-smoothing on the $j$th level.  On the other hand, in Algorithm \ref{alg:fascd} the initial value of the correction $z^j$, before up-smoothing on the $j$th level, is $z^j = y^j + P z^{j-1}$, where $z^{j-1}$ contains the total correction from coarser corrections and up-smoothing at the $j-1$ level.  Thus the $j$th-level iterate \emph{before} up-smoothing must be set to
    $$w^j + y^j + P z^{j-1} = u^j + P(u^{j-1} - \iR u^j).$$

With these modifications, we have a much-simplified pseudocode, Algorithm \ref{alg:fas}.  This FAS V-cycle appears in various places, for example Algorithm 14 of \cite{Bruneetal2015} or in section 5.3.4 of \cite{Trottenbergetal2001}.

>>>>

This paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive variational inequalities, with several example problems, and then Section \ref{sec:cd} reviews the basics of CDs, including the associated iterations, as proposed by Tai \cite{Tai2003}.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, along with the special monotone restriction operators used here.  Section \ref{sec:cdmultilevel} presents defect constraints, leading to an apparently new understanding of telescoping sets for up-smoothing.  Section \ref{sec:vcycle} then presents the FASCD V-cycle, with Section \ref{sec:implementation} addressing implementation and smoother semantics, plus F-cycles.  Section \ref{sec:results} gives FASCD numerical results for the Section \ref{sec:vi} problems, followed by discussion and outlook in Section \ref{sec:discussion}.  An Appendix sketchs how FASCD reduces to classical FAS multigrid for PDEs.

@@

Associated to a constraint decomposition \eqref{eq:subspacedecomp}--\eqref{eq:constraintrestrictionsum} are certain iterative methods \cite{Tai2003,Xu1992}.  Let $0\le \alpha \le 1$ be the \emph{damping parameter}.  Algorithms \ref{alg:basiccd-add} and \ref{alg:basiccd-mult} below, discussed next, solve smaller VI problems over each subset $\cK_i$.  Each Algorithm replaces $u \in \cK$ with a new iterate $w\in\cK$ which should be closer to $u^* \in \cK$ solving \eqref{eq:vi}.

@@

Compared specifically to the $V(1,0)$ cycle presented as Algorithm 4.7 in \cite{GraeserKornhuber2009}, one might say informally that \pr{fascd-vcycle} is a $V(1,1)$ cycle.  However, its structure relates directly to the basic CD iterations (Sections \ref{sec:cd}) as well.  In unilateral cases it is precise to say that one iteration of \pr{fascd-vcycle} is an application of \pr{cd-mult} over CD \eqref{eq:fe:downwardsuminclusion}, using the defect constraint meaning of the downward constraint sets $\mathcal{D}^j$, followed by an application of \pr{icd-tele} (Section \ref{sec:cdmultilevel}) over the incomplete CD \eqref{eq:fe:upwardsuminclusion}.  FIXME: idea is that once you are using defect constraints you no longer depend on the literal availability of $\Pi_i$, so incomplete CD suffices, e.g.~even for \pr{cd-mult}

@@

FIXME F-cycles

\begin{pseudofloat}[ht]
\begin{pseudo}
\pr{fascd-fcycle} \\
FIXME
\end{pseudo}
\caption{The FASCD F-cycle for solving FE VI problem \eqref{eq:fe:vi}.}
\label{alg:fascd-fcycle}
\end{pseudofloat}

@@

@@

FIXME desired extension to nonlocal variational inequalities

@@

\appendix
\section{Reduction to FAS multigrid for PDEs} \label{app:reductions}

\pr{fascd-vcycle} (Algorithm \ref{alg:fascd}) reduces to the classical FAS multigrid V-cycle for nonlinear PDEs \cite{Trottenbergetal2001}, which itself generalizes linear multigrid, when all inequality constraints are removed.  In fact, suppose we remove the inequality constraints from the finest-level VI problem \eqref{eq:fe:vi} by replacing the finest-level obstacles $\underline{\gamma}^J$ and $\overline{\gamma}^J$ with very negative and positive constants, respectively, thus not constraining the solution $u^J$ or any possible iterate.  The admissible set $\mathcal{K}^J$ of \eqref{eq:fe:vi} is then determined only by the Dirichlet boundary conditions.  Let $\mathcal{V}_0^J \subset \mathcal{V}^J$ denote the linear subspace with zero values on $\partial_D\Omega$.  By replacing the (now) arbitrary element $v-u^J\in\mathcal{V}_0^J$ in \eqref{eq:fe:vi} with ``$v$'' we have the finest-level problem
\begin{equation}
\ip{f^J(u^J)}{v} = \ip{\ell^J}{v} \qquad \text{for all } v\in \mathcal{V}_0^J, \label{eq:app:fas:pde}
\end{equation}
the FE discretization of strong-form PDE $f(u)=\ell$.  The downward $\mathcal{D}^j$ and upward $\mathcal{U}^j$ constraint sets can also be enlarged so that the corrections $y^j$ and $z^j$ in \pr{fascd-vcycle} are never constrained.\footnote{Formally we might do this by choosing constant LDCs with $\underline{\chi}^j \ll 0 \ll \overline{\chi}^j$, but so that ordering property \eqref{eq:fe:chiordering} still holds, and also with large gaps so that $\underline{\phi}^j = \underline{\chi}^j - \underline{\chi}^{j-1} \ll 0 \ll \overline{\phi}^j = \overline{\chi}^j - \overline{\chi}^{j-1}$.}  Thus the corrections $y^j$ and $z^j$ are unconstrained, and lines 2--3 and 5--8 can be removed from Algorithm \ref{alg:fascd}.

Now the downward/upward VIs \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi} are also unconstrained, and can be rewritten as weak forms $\ip{f^j(w^j + y^j)}{v} = \ip{\ell^j}{v}$ and $\ip{f^j(w^j + z^j)}{v} = \ip{\ell^j}{v}$, respectively, for all $v\in \mathcal{V}_0^j$.  Then we combine $u^j=w^j+y^j$, instead of $w^j$ and $y^j$ separately, and similarly $u^j=w^j+z^j$ when up-smoothing.  The smoothers and the coarsest-level solver can be regarded as unconstrained, in-place modifications of $u^j$, so we denote them simply as ``$\text{\pr{smooth}}(\ell^j;u^j)$.''  Noting line 11 of Algorithm \ref{alg:fascd} now says $u^{j-1}=\iR u^j$, after down-smoothing the coarsened source functional computed by line 12 can be rewritten as
\begin{equation}
\ell^{j-1} = f^{j-1}\left(u^{j-1}\right) + R\left(\ell^j-f^j(u^j)\right). \label{eq:app:fas:levelsource}
\end{equation}

Going upward we must be careful that only the coarse correction, and not the coarse solution, is prolonged \cite[remark 5.3.9]{Trottenbergetal2001}.  After up-smoothing on the $j-1$ level the coarse iterate is $u^{j-1}$, but at this point $u^j=w^j + y^j$ contains the result of $j$th-level down-smoothing.  The initial value of the correction $z^j$, before up-smoothing on the $j$th level, is $z^j = y^j + P z^{j-1}$, where $z^{j-1}$ contains the total correction from coarser corrections and $j-1$ level up-smoothing.  Thus the $j$th-level iterate \emph{before} up-smoothing must be set to
    $$w^j + y^j + P z^{j-1} = u^j + P(u^{j-1} - \iR u^j).$$

With these modifications, we have a much-simplified Algorithm \ref{alg:fas}.  This FAS V-cycle appears in various places, for example Algorithm 14 of \cite{Bruneetal2015} or in section 5.3.4 of \cite{Trottenbergetal2001}.
